Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-05 paper/paper.tex

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

\paragraph{Concrete interpretation} can be characterized by setting the implicit
parameters of the relation given in Figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}(\ell,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

\begin{align*}
\mathit{push}(\ell,\msto,\mkont) &= \ell,\msto\sqcup[\ell\mapsto\{\mkont\}]
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
\end{align*}

@@

\begin{align*}
\mnum+1 &\in \delta(\saddone,\mnum) &
\mnum-1 &\in \delta(\ssubone,\mnum)\\
\strue &\in \delta(\szerohuh,\szero) &
\sfalse &\in \delta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

\begin{align*}
\sNum &\in \hat\delta(\saddone,\mnum) &
\sNum &\in \hat\delta(\ssubone,\mnum)\\
\strue &\in \hat\delta(\szerohuh,\sNum) &
\sfalse &\in \hat\delta(\szerohuh,\sNum)\\
\strue &\in \hat\delta(\szerohuh,\szero) &
\sfalse &\in \hat\delta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

\begin{figure}
\begin{alltt}
  ;; State \(\rightarrow\) Setof State
  (define (step state)
    (match state
      [(ev σ e ρ k)
       (match e
         [(var\(\superscript\ell\) x)           (for/set ((v (lookup ρ σ x))) (co σ k v))]
         [(lit\(\superscript\ell\) l)           (set (co σ k n))]
         [(lam\(\superscript\ell\) x e)         (set (co σ k (clos x e ρ)))]
         [(app\(\superscript\ell\) f e)
          (define-values (σ* a) (push state))
          (set (ev σ* f ρ (ar e ρ a)))]
         [(ife\(\superscript\ell\) e0 e1 e2)
          (define-values (σ* a) (push state))
          (set (ev σ* e0 ρ (ifk e1 e2 ρ a)))])]
      [(co σ k v)
       (match k
         ['mt (set (ans σ v))]
         [(ar\(\superscript\ell\) e ρ) (set (ev σ e ρ (fn v l)))]
         [(fn\(\superscript\ell\) f)   (for/set ((k (get-cont σ l))) (ap σ f v k))]
         [(fi\(\superscript\ell\) c a ρ)
          (for/set ((k (get-cont σ l)))
            (ev σ (if v c a) ρ k))])]
      [(ap σ fun a k)
       (match fun
         [(clos l x e ρ)
          (define-values (ρ* σ*) (bind state))
          (set (ev σ* e ρ* k))]
         [(? op? o)
          (for*/set ((k (get-cont σ l))
                     (v (\(\Delta\) o (list v))))
            (co σ k v))
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

\begin{alltt}
(define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))

(define pred
  (\(\lambda\) (n)
    (\(\lambda\) (rf)
      (\(\lambda\) (rx)
        (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
          (\(\lambda\) (i) rx))
         (\(\lambda\) (id) id))))))
\end{alltt}

\begin{alltt}
;; multiplication distributes over addition
((church=? ((mult \church2) ((plus \church1) \church3)))
 ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))))
\end{alltt}

@@

;; (State -> Setof State) -> State^ -> { State^ }
(define ((wide-step step) state)
  (match state
    [(cons cs σ)
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c σ))))
     (set (cons (for/set ([s ss]) (s->c s))
                (join-stores ss)))]))
\end{verbatim}

@@

\begin{figure}
\begin{align*}
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Compilation}
\end{figure}

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons σ cs)
     (define-values (cs* σ*)
       (for/fold ([cs* (set)] [σ* σ])
         ([c cs])
         (match (step-compiled^ (cons σ c))
           [(cons σ** cs**)
            (values (set-union cs* cs**) (join-store σ* σ**))])))
     (cons σ* (set-union cs cs*))]))
\end{alltt}
%% Should be commutted to beginning

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons σ cs)
     (define-values (cs* ∆)
       (for/fold ([cs* (set)] [∆* '()])
         ([c cs])
         (match (step-compiled^ (cons σ c))
           [(cons ∆** cs**)
            (values (set-union cs* cs**) (append ∆** ∆*))])))
     (cons (update ∆ σ) (set-union cs cs*))]))
\end{alltt}

@@

We combine the interface of generators ~\cite{ianjohnson:cluhistory}
with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
of {\tt ev} states. We also have a special {\tt do} form that performs
behind-the-scenes state passing and iteration much like {\tt do} in
Haskell. In the presence of the different optimization
techniques, {\tt generator}, {\tt yield}, {\tt $\lambda$\%} and {\tt do}
change meaning slightly, but keep the same interface as their names
suggest.

\begin{figure}
\begin{alltt}
\(e-dispatch =\)
 (match e
    [(var\(\superscript\ell\) x)
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) ((v (delay (lookup-env \(\menv\) x))))
        (yield (co \(\msto\) k v))))]
    [(lit\(\superscript\ell\) l)
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
    [(lam\(\superscript\ell\) x e)
     (define c (compile e))
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
    [(app\(\superscript\ell\) f e)
     (define cf (compile f))
     (define ce (compile e))
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (define-values (\(\msto\)* a) (push state))
      (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
    [(ife\(\superscript\ell\) e0 e1 e2)
      (define c0 (compile e0))
      (define c1 (compile e1))
      (define c2 (compile e2))
      (\(\lambda\)\% (\(\msto\) \(\menv\) k)
        (do (\(\msto\)) [(\(\msto\)* a) (push state)]
         (yield (ev \(\msto\)* c0 \(\menv\) (ifk c1 c2 \(\menv\) a)))))])

  (define (step state)
    (match state
      [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
      [(co \(\msto\) k v)
       (match k
         ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
           (yield (ans \(\msto\) v))))]
         [(ar\(\superscript\ell\) e \(\menv\))
          (generator (do (\(\msto\)) ()
            (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
         [(fn\(\superscript\ell\) f)
          (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
           (yield (ap \(\msto\) f v k))))]
         [(fi\(\superscript\ell\) c a \(\menv\))
          (generator
           (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                    [v <- (force \(\msto\) v)])
             (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
      [(ap \(\msto\) fun a k)
       (match fun
         [(clos l x e \(\menv\))
          (generator
           (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
             (yield (ev \(\msto\)* e \(\menv\)* k))))]
         [(? op? o)
          (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                   [v <- (force \(\msto\) v)]
                   [v (\(\Delta\) o (list v))])
            (yield (co \(\msto\) k v)))
         [_ (generator (do (\(\msto\)) () (continue)))]))]))
\end{alltt}
\caption{Parameterized Abstract$^2$ machine for ISWIM}
\label{fig:paam}
\end{figure}

@@

\bibliographystyle{splncs}
\bibliography{dvh-bibliography}

@@

\begin{verbatim}
 (let ((id (lambda (x) x)))
   (begin (id 1) (id 2)))
\end{verbatim}

>>>>

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

\paragraph{Concrete interpretation} can be characterized by setting the implicit
parameters of the relation given in Figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}(\mlab,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

\begin{align*}
\mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
\end{align*}

@@

\begin{align*}
\mnum+1 &\in \interpdelta(\saddone,\mnum) &
\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
\strue &\in \interpdelta(\szerohuh,\szero) &
\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

\begin{align*}
\sNum &\in \hat\interpdelta(\saddone,\mnum) &
\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
\strue &\in \hat\interpdelta(\szerohuh,\szero) &
\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

HERE: ρ

\begin{figure}
\begin{alltt}
  ;; State \(\rightarrow\) Setof State
  (define (step \(\mstate\))
    (match \(\mstate\)
      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
       (match \(\mexp\)
         [(var\(\superscript\mlab\) \(\mvar\))
          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
      [(co \(\msto\) \(\mkont\) \(\mval\))
       (match \(\mkont\)
         ['mt (set (ans \(\msto\) \(\mval\)))]
         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
         [(fn\(\superscript\mlab\) \(\mval'\))
          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
         [(fi\(\superscript\mlab\) c a \(\rho\))
          (for/set ((k (get-cont \(\msto\) l)))
            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
       (match fun
         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
         [(op \(\mop\))
          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
                     (\(\mval\) (\(\Delta\) \(\mop\) \(\mval\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

%% \begin{alltt}
%% (define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))

%% (define pred
%%   (\(\lambda\) (n)
%%     (\(\lambda\) (rf)
%%       (\(\lambda\) (rx)
%%         (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
%%           (\(\lambda\) (i) rx))
%%          (\(\lambda\) (id) id))))))
%% \end{alltt}

%% (define \church0 (\(\lambda\) (f0) (\(\lambda\) (x0) x0)))
%% (define \church1 (\(\lambda\) (f1) (\(\lambda\) (x1) (f1 x1))))
%% (define \church2 (\(\lambda\) (f2) (\(\lambda\) (x2) (f2 (f2 x2)))))
%% (define \church3 (\(\lambda\) (f3) (\(\lambda\) (x3) (f3 (f3 (f3 x3))))))

\begin{figure}
\begin{alltt}
(define ((((plus p1) p2) pf) x)
  ((p1 pf) ((p2 pf) x)))

(define (((mult m1) m2) mf)
  (m2 (m1 mf)))

(define (((pred n) rf) rx)
  (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
    (\(\lambda\) (ignored) rx))
   (\(\lambda\) (id) id)))

(define (zero? n)
  ((n (\(\lambda\) (zx) #f)) #t))

(define ((church=? e1) e2)
  (if (zero? e1) (zero? e2)
      (if (zero? e2) #f
          ((church=? (pred e1)) (pred e2)))))

;; multiplication distributes over addition
((church=? ((mult \church2) ((plus \church1) \church3)))
 ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
\end{alltt}
\caption{Church-numeral calculation}
\end{figure}

@@

;; (State -> Setof State) -> State^ -> { State^ }
(define ((wide-step step) state)
  (match state
    [(cons cs \(\msto\))
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c \(\msto\)))))
     (set (cons (for/set ([s ss]) (s->c s))
                (join-stores ss)))]))
\end{verbatim}

@@

\begin{figure}
\begin{align*}
\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
\\
\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\mlit,\msto}
\\
\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
\\
\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
\\
&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
\\
&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\end{align*}
\caption{Compilation}
\end{figure}

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* \(\msto\)*)
       (for/fold ([cs* (set)] [\(\msto\)* \(\msto\)])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons \(\msto\)** cs**)
            (values (set-union cs* cs**) (join-store \(\msto\)* \(\msto\)**))])))
     (cons \(\msto\)* (set-union cs cs*))]))
\end{alltt}
%% Should be commutted to beginning

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* ∆)
       (for/fold ([cs* (set)] [∆* '()])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons ∆** cs**)
            (values (set-union cs* cs**) (append ∆** ∆*))])))
     (cons (update ∆ \(\msto\)) (set-union cs cs*))]))
\end{alltt}

@@

We combine the interface of generators~\cite{ianjohnson:cluhistory}
with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
of {\tt ev} states. We also have a special {\tt do} form that performs
behind-the-scenes state passing and iteration much like {\tt do} in
Haskell. In the presence of the different optimization
techniques, {\tt generator}, {\tt yield}, {\tt $\lambda$\%} and {\tt do}
change meaning slightly, but keep the same interface as their names
suggest.

\begin{figure}
\begin{alltt}
\(e-dispatch =\)
 (match e
    [(var\(\superscript\mlab\) x)
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) ((v (delay (lookup-env \(\menv\) x))))
        (yield (co \(\msto\) k v))))]
    [(lit\(\superscript\mlab\) l)
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
    [(lam\(\superscript\mlab\) x e)
     (define c (compile e))
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
    [(app\(\superscript\mlab\) f e)
     (define cf (compile f))
     (define ce (compile e))
     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
      (define-values (\(\msto\)* a) (push state))
      (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
    [(ife\(\superscript\mlab\) e0 e1 e2)
      (define c0 (compile e0))
      (define c1 (compile e1))
      (define c2 (compile e2))
      (\(\lambda\)\% (\(\msto\) \(\menv\) k)
        (do (\(\msto\)) [(\(\msto\)* a) (push state)]
         (yield (ev \(\msto\)* c0 \(\menv\) (ifk c1 c2 \(\menv\) a)))))])

  (define (step state)
    (match state
      [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
      [(co \(\msto\) k v)
       (match k
         ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
           (yield (ans \(\msto\) v))))]
         [(ar\(\superscript\mlab\) e \(\menv\))
          (generator (do (\(\msto\)) ()
            (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
         [(fn\(\superscript\mlab\) f)
          (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
           (yield (ap \(\msto\) f v k))))]
         [(fi\(\superscript\mlab\) c a \(\menv\))
          (generator
           (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                    [v <- (force \(\msto\) v)])
             (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
      [(ap \(\msto\) fun a k)
       (match fun
         [(clos l x e \(\menv\))
          (generator
           (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
             (yield (ev \(\msto\)* e \(\menv\)* k))))]
         [(? op? o)
          (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                   [v <- (force \(\msto\) v)]
                   [v (\(\Delta\) o (list v))])
            (yield (co \(\msto\) k v)))
         [_ (generator (do (\(\msto\)) () (continue)))]))]))
\end{alltt}
\caption{Parameterized Abstract$^2$ machine for ISWIM}
\label{fig:paam}
\end{figure}

@@

\bibliographystyle{splncs}
\bibliography{bibliography}

@@

\begin{verbatim}
 (let ((id (\(\lambda\) (x) x)))
   (begin (id 1) (id 2)))
\end{verbatim}
